---
title: "CDS 101 – Final Project Report"
author: "Samir Dawar, Andrew Lee, Sumera"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---


# 1. Problem Definition
The Chernobyl nuclear disaster released radioactive materials into the atmosphere, and these contaminants were carried across Europe by wind and weather systems. In this project, we study how the concentrations of three isotopes (Cs-137, Cs-134, and I-131) vary with distance from the Chernobyl Nuclear Power Plant. Our main goal is to see whether radiation levels decrease as distance increases and whether each isotope behaves differently due to differences in half-life. Cs-137 has a long half-life (about 30 years), Cs-134 decays faster, and I-131 decays very quickly (around 8 days). By comparing them, we can understand which isotopes show clear geographic decay and which are more influenced by timing and environmental conditions.



# 2. Data Acquisition & Description
We used the “Chernobyl Chemical Radiation / CSV / Country Data” dataset from Kaggle. After downloading the file, we saved it in our project folder under `data/Chernobyl_Chemical_Radiation.csv`.

The dataset contains:
- Country code (`PAYS`)
- Monitoring station name (`Location`)
- Longitude and latitude
- Date of measurement
- Concentrations of I-131, Cs-134, and Cs-137 (Bq/m3)

Each row is one observation from a monitoring location on a specific day shortly after the accident. The dataset does not include any personal data, so there are no privacy concerns.

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

# 3. Data Cleaning & Preprocessing
We selected the variables needed for our analysis and renamed the isotope columns to shorter and easier names. Rows with missing isotope values were removed. To study how radiation changes with distance, we created a new variable called `distance` using a simple distance formula based on longitude and latitude differences.

```{r cleaning, warning=FALSE}
chernobyl <- read.csv("data/Chernobyl_ Chemical_Radiation.csv")

chernobyl_clean <- chernobyl %>%
  select(
    PAYS,
    Location,
    Longitude,
    Latitude,
    I_131 = I_131_.Bq.m3.,
    Cs_134 = Cs_134_.Bq.m3.,
    Cs_137 = Cs_137_.Bq.m3.
  ) %>%
  mutate(
    I_131 = as.numeric(I_131),
    Cs_134 = as.numeric(Cs_134),
    Cs_137 = as.numeric(Cs_137)
  ) %>%
  filter(!is.na(Cs_137))

#distrance measurement
chernobyl_clean <- chernobyl_clean %>%
  mutate(
    distance = sqrt((Longitude - 30.099)^2 + (Latitude - 51.389)^2)
  )
head(chernobyl_clean)
  
```

# 4. Exploratory Data Analysis (EDA)
The goal of this section is to explore the dataset and understand how radiation levels differ across countries and how they change with distance from the Chernobyl plant. We look at summary statistics, boxplots, scatterplots, and a geographic-style plot to visually assess the relationships that will later be modeled. We begin by examining simple descriptive statistics for the three isotopes.

```{r, warning=FALSE}
summary(chernobyl_clean[, c("I_131", "Cs_134", "Cs_137")])
```
The summary statistics show that all three isotopes have very skewed distributions. Most values are close to zero, but there are a few large measurements, especially for Cs-137 and I-131. This is expected because only certain monitoring stations recorded heavy fallout. Cs-134 has more missing values, which reflects gaps in collection across countries. These summaries help us understand the starting point before plotting.

```{r, warning=FALSE}
ggplot(chernobyl_clean, aes(x = PAYS, y = Cs_137)) +
geom_boxplot() +
labs(
title = "Cs-137 Levels by Country",
x = "Country",
y = "Cs-137 (Bq/m3)"
)
```
This boxplot shows that Cs-137 levels vary noticeably across different countries. Some countries have higher medians and more extreme outliers, while others have consistently low measurements. Because Cs-137 is a long-lasting isotope, these differences likely reflect how fallout patterns moved across Europe after the explosion. This supports the idea that geographic location influenced radiation exposure levels.

```{r, warning=FALSE}
ggplot(chernobyl_clean, aes(distance, Cs_137)) +
geom_point(alpha = 0.6) +
labs(
title = "Cs-137 vs Distance from Chernobyl",
x = "Distance",
y = "Cs-137 (Bq/m3)"
)
```
Cs-137 clearly decreases as distance from Chernobyl increases. Locations closer to the plant have higher concentrations and more extreme values, while farther locations cluster near zero. This matches scientific expectations because Cs-137 persists in the environment and spreads outward with decreasing intensity. This visualization directly supports our main research question.

```{r, warning=FALSE}
ggplot(chernobyl_clean, aes(distance, Cs_134)) +
geom_point(alpha = 0.6) +
labs(
title = "Cs-134 vs Distance from Chernobyl",
x = "Distance",
y = "Cs-134 (Bq/m3)"
)
```
Cs-134 also shows a downward pattern with distance, but the trend is weaker than Cs-137. Many values are close to zero, and there is more variation overall. This is expected because Cs-134 decays faster than Cs-137, so its measurements depend more on timing and local environmental conditions. Still, the overall pattern suggests that distance plays some role in Cs-134 concentration.

```{r, warning=FALSE}
ggplot(chernobyl_clean, aes(distance, I_131)) +
geom_point(alpha = 0.6) +
labs(
title = "I-131 vs Distance from Chernobyl",
x = "Distance",
y = "I-131 (Bq/m3)"
)
```
I-131 does not show a clear relationship with distance. Measurements are scattered at many distances, including some high values that appear far from the plant. This behavior makes sense because I-131 has a very short half-life (about eight days), so its concentration depends more on the date of measurement and weather patterns than on geographic distance. Distance alone cannot explain I-131 levels.

```{r, warning=FALSE}
ggplot(chernobyl_clean, aes(Longitude, Latitude, color = Cs_137)) +
geom_point() +
labs(
title = "Geographic Distribution of Cs-137",
x = "Longitude",
y = "Latitude",
color = "Cs-137"
)
```
This map-style plot shows where the highest Cs-137 values occurred across Europe. Higher concentrations appear in areas closer to the Chernobyl plant and in regions where prevailing winds carried the fallout. This plot reinforces that contamination was not uniform across Europe and supports the idea that both distance and location influenced radiation levels.

# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- Explain why your plot types are appropriate.  
- Comment on labels, legends, colors, and overall readability.  
- Mention any steps you took to make plots accessible and interpretable.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
