---
title: "CDS 101 – Final Project Report"
author: "Samir Dawar, Andrew Lee, Sumera"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---


# 1. Problem Definition
The Chernobyl nuclear disaster released radioactive materials into the atmosphere, and these contaminants were carried across Europe by wind and weather systems. In this project, we study how the concentrations of three isotopes (Cs-137, Cs-134, and I-131) vary with distance from the Chernobyl Nuclear Power Plant. Our main goal is to see whether radiation levels decrease as distance increases and whether each isotope behaves differently due to differences in half-life. Cs-137 has a long half-life (about 30 years), Cs-134 decays faster, and I-131 decays very quickly (around 8 days). By comparing them, we can understand which isotopes show clear geographic decay and which are more influenced by timing and environmental conditions.



# 2. Data Acquisition & Description
We used the “Chernobyl Chemical Radiation / CSV / Country Data” dataset from Kaggle. After downloading the file, we saved it in our project folder under `data/Chernobyl_Chemical_Radiation.csv`.

The dataset contains:
- Country code (`PAYS`)
- Monitoring station name (`Location`)
- Longitude and latitude
- Date of measurement
- Concentrations of I-131, Cs-134, and Cs-137 (Bq/m3)

Each row is one observation from a monitoring location on a specific day shortly after the accident. The dataset does not include any personal data, so there are no privacy concerns.

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

# 3. Data Cleaning & Preprocessing
We selected the variables needed for our analysis and renamed the isotope columns to shorter and easier names. Rows with missing isotope values were removed. To study how radiation changes with distance, we created a new variable called `distance` using a simple distance formula based on longitude and latitude differences.

```{r cleaning, eval=FALSE}
chernobyl <- read.csv("data/Chernobyl_Chemical_Radiation.csv")
chernobyl_clean <- chernobyl %>%
  
```

# 4. Exploratory Data Analysis (EDA)

This section maps directly to the **EDA** part of the rubric:

- Summary statistics  
- Visualizations  
- Interpretation connected to the research question  

```{r eda-summary, message=FALSE, warning=FALSE}
# library(dplyr)
# summary(data_clean)
```

```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Example histogram of a numeric variable (EDA rubric)."}
# library(ggplot2)
# ggplot(data_clean, aes(x = some_numeric_variable)) +
#   geom_histogram(binwidth = 5, color = "white") +
#   labs(x = "Some Variable", y = "Count", title = "Distribution of Some Variable")
```

# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- Explain why your plot types are appropriate.  
- Comment on labels, legends, colors, and overall readability.  
- Mention any steps you took to make plots accessible and interpretable.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
